---
title: "Lab 6: R and SQL"
author: "Alex Ptacek"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

# Overview

This lab is divided into two parts. In the first part you will practice using joins for data 
wrangling and analysis on the `nycflights` dataset. Some of the problems come from Chapter 19 of your book. For the second part, you will download a dataset on the budgets of college sports programs and process it for storage in a relational database (I strongly recommend using `duckdb` which can be installed using `install.packages("duckdb")`- duckdb is highly performant, self-contained, and ideally suited both to learning SQL and performing data analysis). Then you will load this database and use `dbplyr` to perform an analysis. You will also practice using `forcats` to recode some of the variables as factors (which are supported by duckdb) and using `separate_wider_delim` to split columns
of text data.

You will need to have installed and to the following libraries:
```{r}
#| echo: true
#| warning: false
library(tidyverse)
library(DBI)
library(duckdb)
library(nycflights13)

```

# Problems

**Part I: Airline Flight Delays**

For the first part of this lab exercise, we will be using the `nycflights` library, which contains 
several different built in datasets including `planes`, which has information on each plane that
appears in the data; `flights`, which has information on individual flights; `airports`, which has information on individual airports; and `weather`, which has information on the weather at the origin airports. In order to do this set of lab exercises, you will need to use different types of joins to combine variables in each data frame.

**Problem 1**

- Use the `flights` and `planes` tables to compute the mean departure delay of each aircraft that has more than 30 recorded flights in the dataset. Hint: Make note of the fact that the variable `year` appears in both `flights` and `planes` but means different things in each before performing any joins.

```{r}
flights |> 
  left_join(planes, by = join_by(tailnum)) |> 
  group_by(model) |> 
  summarise(n = n(), mean_dep_delay = mean(dep_delay)) |> 
  filter(n >30)
```


- Use `anti-join` to identify flights where `tailnum` does not have a match in `plane`. Determine
the carriers for which this problem is the most common. 

```{r}
flights |> 
  anti_join(planes, by = join_by(tailnum)) |> 
  count(carrier) |> 
  arrange(desc(n)) |> 
  left_join(airlines) |> 
  relocate(name, .before = n)
```


- Find the airplane model which made the most flights in the dataset, and filter the dataset to contain only flights flown by airplanes of that model, adding a variable which corresponds to the year each those airplanes were built. Then compute the average departure delay for each year of origin and plot the data. Is there any evidence that older planes have more greater departure delays?

*Answer: Older manufactures of the airplane model with the most flights seem to have greater departure delays.*

```{r most-flights-model-avg-dep-delay, warning=FALSE}
#Find model with most flights
most_flights_model <- flights |> 
  left_join(planes, by = join_by(tailnum)) |> 
  count(model) |> 
  slice_max(order_by = n, n = 2) |> 
  filter(model != is_null(model)) #Filtering out the null value as this likely represents multiple models

#Find average dep_delay by year_built for this model
flights |> 
  left_join(planes, by = join_by(tailnum)) |> 
  rename(year_built = year.y) |> 
  filter(model %in% most_flights_model$model) |> 
  group_by(year_built) |> 
  summarise(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) |> 
  
#Plot this table
  ggplot(aes(x = year_built, y = mean_dep_delay)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "salmon")
```


**Problem 2**

- Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:

```{r}
#| eval: false
#| echo: true
airports |>
  semi_join(flights, join_by(faa == dest)) |>
  ggplot(aes(x = lon, y = lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()

```
You might want to use the size or color of the points to display the average delay for each airport.

```{r}
flights |> 
  group_by(dest) |> 
  summarise(mean_arr_delay = mean(arr_delay)) |> 
  left_join(airports, join_by(dest == faa)) |> 
  ggplot(aes(x = lon, y = lat, size = mean_arr_delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()
```


**Part II: Creating and Accessing a Database **

In this exercise we will begin with a flat file which contains data on college sports programs throughout the country. The source of the data is a government run database called [Equity in Athletics Data Analysis](https://ope.ed.gov/athletics/#/), though we are working with just
a small subset here. You can download this file by clicking here:
[sports_program_costs.csv](https://github.com/georgehagstrom/DATA607/tree/main/website/assignments/labs/labData/sports_program_costs.csv). I have also included a data dictionary which gives a quick description of the dataset, which can be downloaded from here: [sports_program_data_dictionary.qmd](https://github.com/georgehagstrom/DATA607/tree/main/website/assignments/labs/labData/sports_program_data_dictionary.qmd). This file contains information on two types of entities: sports teams and universities, however the information on both entities is combined into a single table, creating substantial redundancies. This exercise has several goals:

1. Load this data into R, split the dataframe into two dataframes, one corresponding to colleges and another corresponding to sports teams, related to each other by common keys. Many databases are stored according to normalization rules, which are designed to limit redundancy and to make it easier to both work with the data and make changes to it. By splitting the data frame we will partially normalize it (but won't go too far).
2. Create a relational database using `duckdb` which contains these two tables.
3. Read this database into R and a/an SQL query/queries to perform an analysis. 

**Problem 3:**

- `sports_program_data.csv` contains variables which either describe properties of a sports team or a college. Split `sports_programs_data` into two data frames, one called `colleges`
and another called `teams`. How can you tell which variables describe colleges and which describe teams? Use the data dictionary and observations of how the values vary as you move from college to college to help make the decision easier. Make sure there are primary keys for both the colleges and teams data frames (verify with `count`)- what are the primary keys in each case and are they simple keys (one variable) or compound keys (require multiple variables). One of these data-sets should contain a foreign key- which one has it and what variables comprise it?

```{r}
sports_program_costs <- read_csv("https://raw.githubusercontent.com/georgehagstrom/DATA607/refs/heads/main/website/assignments/labs/labData/sports_program_costs.csv")
```

*Answer: I could tell which data belonged to `colleges` and which belonged to `teams` based on the data dictionary. The `colleges` data has variables that only identify aspects of the school. This includes the 3 sports division columns, such as classification code, because each school can only be apart of 1 sports division. It does not include the granularity of individual sport-level data, such as sport, gender makeup, and revenue. The `teams` data includes individual sport-level data (i.e. the remaining columns), plus the school code (`unitid`) to identify the team, and the `year` because observations are yearly. The primary key for `colleges` is `year` and `unitid`, and the primary key for `teams` is `year`, `unitid`, and `sportscode`. These are both compound keys. The foreign key is in `teams` and it is `year` and `unitid`.*

```{r}
#Create colleges table
colleges <- sports_program_costs |> 
  select(year:sector_name) |> 
  group_by(across(year:classification_other), sector_cd, sector_name) |> 
  summarise(ef_male_count = sum(ef_male_count),
            ef_female_count = sum(ef_female_count),
            ef_total_count = sum(ef_total_count))

#Check colleges primary key
colleges |> 
  count(year, unitid) |> 
  filter(n > 1)

#Create teams table
teams <- sports_program_costs |> 
  select(year, unitid, sportscode:sports) 

#Check teams primary key
teams |> count(year, unitid, sportscode) |> 
  filter(n >1)
```


- The variable `sector_name` contains information about whether a college is public, private, 
non-profit, for-profit, a 2-year college, or a 4-year + college. Split this variable (using `separate_wider_delim`) into two variables, one of which describes whether the college is a Public, Private nonprofit, or private for-profit, and another which describes how many years the college programs run.

```{r}
#Examine sector_name variable
view_sect_names <- colleges |> 
  group_by(sector_name) |> 
  count(sector_name)
view_sect_names |> head(n = 5)

#Separate sector_name by comma as delim
colleges <- colleges |> 
  separate_wider_delim(cols = sector_name, delim = ", ", 
                       names = c("sector_type", "degree_type"),
                       cols_remove = FALSE) 

#Confirm changes worked
view_sect_names2 <- colleges |> 
  group_by(sector_type, degree_type) |> 
  count(sector_type, degree_type)
view_sect_names2 |> head(n = 5)
```


- Several variables are candidates to be recoded as factors, for example `state_cd`, 
`zip_text`, `classification_name`, `sports`, and the `sector` variables you just created for the previous part. Recode these variables as categorical variables. For the `classification` variable, use the `classification_code` to order the factors according to the numeric code. 

```{r}
#Create state_order as levels for state_cd
state_order <- append(state.abb, "DC", after = 7)
state_order <- append(state_order, "PR", after = 38)
state_order <- append(state_order, "VI", after = 47)

#Mutate state_cd into factor
colleges <- colleges |> 
  mutate(state_cd = fct(state_cd, levels = state_order))


#Create sorted_zips as levels for zip_text
sorted_zips <- fct_reorder(colleges$zip_text, as.integer(colleges$zip_text), .na_rm = FALSE)

#Mutate zip_text into factor
colleges <- colleges |> 
  mutate(zip_text = fct(zip_text, levels = levels(sorted_zips)))


#Create sorted_class as levels for classification_name
class_order <- fct_reorder(colleges$classification_name, colleges$classification_code)

#Mutate classification_name into factor
colleges <- colleges |> 
  mutate(classification_name = fct(classification_name, levels = levels(class_order)))


#Create sports_order as levels for sports
sports_order <- fct_reorder(teams$sports, teams$sportscode)

#Mutate sports into factor
teams <- teams |> 
  mutate(sports = fct(sports, levels = levels(sports_order)))


#Create sector_order as levels for sector_name
sector_order <- fct_reorder(sports_program_costs$sector_name, sports_program_costs$sector_cd)

#Mutate sector_name into factor
colleges <- colleges |> 
  mutate(sector_name = fct(sector_name, levels = levels(sector_order)))
  

#Observe changes
glimpse(colleges)
glimpse(teams)
```


**Problem 4** 

- Using `DBI`, `duckdb`, and `dbplyr`, create a relational database with two tables, writing the `sports` data frame you created in problem 3 to one and the `colleges` data frame (also from problem 3) to the other. Write this database to disk. How does the size of the database file compare to the original csv?

*Answer: The original file was 24.4MB. The new relational database is 4.5MB.*

```{r }
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = "duckdb")

dbWriteTable(con, "colleges", colleges)
dbWriteTable(con, "teams", teams)

dbListTables(con)
```


- Use `dbplyr` to write a query to this database that calculates the top 10 colleges ranked by
the average profit (defined as revenue - expenses) of their american football team over the years of data. Print the SQL query that results from your R pipeline using `show_query()` and then use `collect()` to show the results of this query.

```{r warning=FALSE}
#Load dbplyr tables
colleges_tbl <- tbl(con, "colleges")
teams_tbl <- tbl(con, "teams")

teams_tbl <-  teams_tbl |> 
  mutate(profit = total_rev_menwomen - total_exp_menwomen)

colleges_tbl |> 
  left_join(teams_tbl, by = join_by(year, unitid)) |> 
  filter(sportscode == 7) |> 
  group_by(institution_name) |> 
  summarise(mean_profit = mean(profit)) |> 
  arrange(desc(mean_profit)) |> 
  head(10) |> 
  show_query() |> 
  collect()
```






