---
title: "Lab 9: Graphs and Social Networks in Pride & Prejudice"
author: "Alex Ptacek"
format: pdf
editor_options: 
  chunk_output_type: console
---

## Overview

In the this lab assignment, you are going to construct a social network 
from the characters in the book "Pride & Prejudice", a novel written by Jane Austen
and available in the `janeaustenr` package. The social network will be a weighted
graph connecting the characters, where the weight is equal to the number of times the
names of each character appeared in each 10 line section of the book. Once you create the graph, you
will load it into tidygraph, make a visualization of the graph, and rank the most 
connected characters by a measure called `degree centrality`

**Problem 1**

Load the text of Pride & Prejudice into R using the `janeaustenr` library. Then download and read [pride_prejudice_characters.csv](https://github.com/georgehagstrom/DATA607/blob/main/website/assignments/labs/labData/pride_prejudice_characters.csv), the
csv file from my github page containing a list of characters in `Pride & Prejudice` and their aliases. 
Here aliases refers to the different names that the characters go by in the books, for example "Darcy"
also goes by the names "Mr. Darcy", and "Mr. Fitzwilliam Darcy" (not to be confused with his cousin "Colonel Fitzwilliam"). 

Process the text of Pride & Prejudice to replace instances where an 
alias occurs with the full name of the character- I recommend using the iteration techniques you learned ealier, I arranged the order of names in the `csv` file to minimize misidentifications if you replace names in the order that they appear in the file. Making this perfect would require a bit of
effort but we are ok if there are some misidentifications. Here the final name of each character will
be a single word. 

```{r load-pkg, message=FALSE}
library(tidyverse)
library(tidytext)
library(janeaustenr)
library(widyr)
library(tidygraph)
library(ggraph)
```

```{r message=FALSE}
base_text <- austen_books() |>
                      filter(book == "Pride & Prejudice")
pride_and_prej_text <- austen_books() |>
                      filter(book == "Pride & Prejudice")

pp_chars <- read_csv("https://raw.githubusercontent.com/georgehagstrom/DATA607/refs/heads/main/website/assignments/labs/labData/pride_prejudice_characters.csv") 


adj_chars <- pp_chars |> 
  mutate(check = ifelse(unique_name == alias, TRUE, FALSE)) |> 
  filter(check == FALSE) |> 
  select(1:3)


fun_chars <- str_flatten(adj_chars$alias, "|") 
fun_chars <- str_c('"', "^", fun_chars, "$", '"')



text_match_df <- tibble(x = ifelse(str_detect(pride_and_prej_text$text, "Jane Austen") == TRUE,
                             pride_and_prej_text$text,
                             str_match_all(pride_and_prej_text$text, fun_chars))) |> 
                  mutate(x = as.character(x),
                         x = str_remove_all(x, "c\\(|\\)"))



text_match_df <- text_match_df |> 
  separate_wider_delim(x, ", ", names = c("x1", "x2", "x3", "x4"), too_few = "align_start" ) |> 
  mutate(across(x1:x4, \(x) str_remove_all(x, '"'))) |> 
  left_join(pp_chars, by = join_by(x1 == alias)) |> 
  left_join(pp_chars, by = join_by(x2 == alias)) |> 
  left_join(pp_chars, by = join_by(x3 == alias)) |> 
  left_join(pp_chars, by = join_by(x4 == alias)) |> 
  select(6,8,10,12) |> 
  rename_at(vars(unique_name.x:unique_name.y.y), ~c("one", "two", "three", "four")) 



pride_and_prej_text <- pride_and_prej_text |> 
  mutate(text = ifelse(str_detect(pride_and_prej_text$text, "Jane Austen") == TRUE,
                       pride_and_prej_text$text, 
                       str_replace(pride_and_prej_text$text, fun_chars, text_match_df$one)))


  

```


**Problem 2**

Following the example in chapter 4 of the text mining with R book, create a new column in the data frame corresponding to the Pride & Prejudice text that divides the text into sections of 10 lines each. Then use the `pairwise_count` function from `widyr` to determine the number of times each name occurs
with each other name in the same 10 line section. 

```{r}
ten_lines_text <- pride_and_prej_text |> 
  mutate(section = row_number() %/% 10) |> 
  filter(section > 0) |> 
  unnest_tokens(word, text) |> 
  filter(!word %in% stop_words$word)

word_pairs <- ten_lines_text |> 
  pairwise_count(word, section, sort = TRUE, upper = FALSE)


```


**Problem 3**

Create a dataframe of nodes which contains the id and unique names of each character, and create a 
dataframe of edges which contains three columns: a column named `from`, a column named `to`, and a column named `weight`, where the `from` and `to` are the id numbers of each character and `weight` is the number of co-occurences you found in Problem 2. Each pair should only appear once in the edge list (i.e. Elizabeth and MrDarcy but not MrDarcy and then Elizabeth). Create a tidygraph object using `tbl_graph` that contains the social network data that we just constructed.

```{r}

adj_chars <- adj_chars |> 
  mutate(unique_name = str_to_lower(unique_name))

word_pairs <-  word_pairs |> 
  filter(item1 %in% adj_chars$unique_name,
         item2 %in% adj_chars$unique_name)

word_pairs <- word_pairs |> 
  rename(from = item1,
         to = item2,
         weight = n)

name_graph <- as_tbl_graph(word_pairs)

name_graph <- name_graph |> 
  activate(nodes) |> 
  mutate(title = str_to_title(name),
         label = str_replace_all(title, " ", "\n"))

```



**Problem 4**

Using `ggraph`, graph the connections between the characters. Make sure that each node is labeled
by the character name, and make sure that the weight is represented by the thickness of the edge
plotted between the two nodes. Then use the `centrality_degree` function to calculate the weighted degree centrality of each character, and make a plot which shows the degree centrality of each character where the characters are arranged in order of degree centrality. 


```{r}
thm <- theme_minimal() +
  theme(
    legend.position = "none",
     axis.title = element_blank(),
     axis.text = element_blank(),
     panel.grid = element_blank(),
     panel.grid.major = element_blank(),
  ) 
theme_set(thm)


name_graph |> 
  ggraph(layout = "kk") +
    geom_node_text(aes(label = label, color = name), size = 3) +
    geom_edge_diagonal(aes(width = weight), color = "gray", alpha = .4)

name_graph |> 
  activate(nodes) |> 
  mutate(deg_centrality = centrality_degree()) |> 
  ggraph(layout = "kk") +
    geom_node_text(aes(label = label, color = name, size = deg_centrality)) +
    geom_edge_diagonal(color = "gray", alpha = 0.4)
```








